---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
```


# shinyEventLogger

[![Licence](https://img.shields.io/badge/licence-MIT-blue.svg)](https://www.r-project.org/Licenses/MIT) 
[![Lifecycle](https://img.shields.io/badge/lifecycle-experimental-organge.svg)](https://www.tidyverse.org/lifecycle/)
[![CRAN Status](https://www.r-pkg.org/badges/version-ago/shinyEventLogger)](https://cran.r-project.org/package=shinyEventLogger)
[![CRAN Checks](https://cranchecks.info/badges/summary/shinyEventLogger)](https://cran.r-project.org/web/checks/check_results_shinyEventLogger.html)
[![Monthly downloads badge](http://cranlogs.r-pkg.org/badges/last-month/shinyEventLogger)](https://cran.r-project.org/package=shinyEventLogger)
[![Daily downloads badge](https://cranlogs.r-pkg.org/badges/last-day/shinyEventLogger?color=blue)](https://CRAN.R-project.org/package=shinyEventLogger)
[![Weekly downloads badge](https://cranlogs.r-pkg.org/badges/last-week/shinyEventLogger?color=blue)](https://CRAN.R-project.org/package=shinyEventLogger)
[![HitCount](http://hits.dwyl.io/kalimu/shinyEventLogger.svg)](http://hits.dwyl.io/kalimu/shinyEventLogger)




by [Kamil Wais](https://kalimu.github.io/)

> **Logging Events in Complex Shiny Apps**. 
    Logging framework dedicated for complex shiny apps. 
    Different types of events can be logged (character string, 
    value of a variable, multi-line output of a function,
    result of a unit test, custom error, warning, or diagnostic message).
    Each event can be logged with a list of parameters that are event-specific, 
    common for a group of events, or common for all app events.
    Logging can be done simultaneously to R console, 
    browser JavaScript console, a file log, and a database (currently MongoDB).
    Log data can be further analyzed with the help of process-mining techniques 
    from 'bupaR' package.


## Installation

### Installing stable version from CRAN

```{r, echo=TRUE, eval=FALSE, cache=FALSE}
install.packages("shinyEventLogger")
```

### Installing version in development from GitHub

```{r, echo=TRUE, eval=FALSE, cache=FALSE}
# install.packages('devtools')
devtools::install_github("kalimu/shinyEventLogger")
```

## Demo Apps

Simple app logging different events to R console, 
browser JavaScript console and to a file.

```{r, echo=TRUE, eval=FALSE, cache=FALSE}
shinyEventLogger::run_demo()
```

Dashboard that allows for interactive analysis of events from demo app.

```{r, echo=TRUE, eval=FALSE, cache=FALSE}
shinyEventLogger::run_demo_dashboard()
```

## Hello World

```{r, echo=TRUE, eval=FALSE, cache=FALSE}
library(shiny)
library(shinyEventLogger)
shinyApp(
  ui = fluidPage(log_init()),
  server = function(input, output) {
    set_logging()
    log_event("Hello World")
 }
)
```

Executing the code above you should see in the console something like this:

    Logging to R console:          ENABLED
    Logging to JavaScript console: ENABLED
    Logging to a file:             disabled
    Logging to a database:         disabled

    |#1|EVENT|Hello World|FIRED|

## Documentation

Project description: [https://kalimu.github.io/shinyEventLogger/](https://kalimu.github.io/shinyEventLogger/)

Package pkgdown homepage:

Vignettes:
* TODO







