---
title: "Introduction to Logging Events in Shiny Apps"
author: "Kamil Wais"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Logging Events in Shiny Apps}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  eval = FALSE,
  cache = FALSE
)
```

> R package `shinyEventLogger` is a logging framework especially usefull
for developing, testing and debugging complex shiny apps.

## Start logging

To be able to use logging inside your shiny app you need to load
both `shiny` and `shinyEventLogger` packages.

```{r}
library(shiny)
library(shinyEventLogger)
```

### Logging to R console

Before logging anything you need to use `set_logging()` to tell the logger
where you want to log events from your app. The `set_logging()` function
should be used before any `log_*` function calls, possibly at the beginning 
of the shiny `server` function.

```{r}
shinyApp(
  ui = fluidPage(),
  server = function(input, output) {
    set_logging()
 }
)
```

You should be able to see in the R console the following messages:

    Logging to R console:          ENABLED
    Logging to JavaScript console: ENABLED
    Logging to a file:             disabled
    Logging to a database:         disabled

Logging to R console is enabled so you can log your first event.

```{r}
shinyApp(
  ui = fluidPage(),
  server = function(input, output) {
    set_logging()
    log_event("Hello World!")
 }
)
```

In the R console you should be able to see something like this:

    Listening on http://127.0.0.1:7082
    Logging to R console:          ENABLED
    Logging to JavaScript console: ENABLED
    Logging to a file:             disabled
    Logging to a database:         disabled

    |#1|EVENT|Hello World!|FIRED|

### Logging to JS console

By default logging will be done to both the R console and 
to browser JavaScript (JS) console. 
Logging to JS console needs also registering some JavaScript code, 
which can be done with the `log_init()` inside `ui` function.

```{r}
shinyApp(
  ui = fluidPage(
    log_init()
  ),
  server = function(input, output) {
    set_logging()
    log_event("Hello World!")
    log_event("See browser JavaScript console (CTRL + SHIFT + I)")
 }
)
```

Now, you should be able to see your logged events both in the R console and
the browser JS console which can be opened by (CTRL + SHIFT + I) shortcut.


## Event structure

Each event instance has a name, type, and status defined by a user. 
It could also have some additional parameters which are
logged together with the event as a list.
Each event has also an event counter - a number that increments every time
any event is fired in the given session. Using `set_logging()` resets the counter.

Let's log an event with some unusual parameters.

```{r}
shinyApp(
  ui = fluidPage(log_init()),
  server = function(input, output) {
    set_logging()
    log_event("Event body", 
              name   = "Event name",
              type   = "NEWTYPE", 
              status = "NEWSTATUS", 
              params = list("param_1" = "param_1_value"))
 }
)
```

In R & JS consoles the result you are able to see is a multi-line output 
with event meta-data in the first line, 
event parameters in the second line,
and event body in the following line.
All lines have the same event counter at the beginning.

    |#1|NEWTYPE|Event name|NEWSTATUS|
    |#1|PARAMS|list(param_1 = "param_1_value")
    |#1|Event body


## Logging different event types

There are several pre-defined event types. 
Each type has its own wrapper function for easier and faster access.
Currently, the following logging function are available:

* `log_event()`
* `log_value()`
* `log_output()`
* `log_test()`,
* `log_message()`, `log_warning()`, and `log_error()`
* `log_started()` and `log_done()`.

### Generic events 

 `log_event()` is a generic function for logging all kinds of events. Default event type is `EVENT` and event status is `FIRED`.
 
If there is no event name defined, objects passed to `...` argument
are evaluated nad coerced into event name.

```{r}
shinyApp(
  ui = fluidPage(log_init()),
  server = function(input, output) {
    set_logging()
    log_event("Event body", name   = "Event name")
    log_event("Event without name")
    log_event(paste("Event", "without", "name"))
    log_event("Event", "without", "name")
 }
)
```

So events 2, 3, and 4 are logged with the same event name, 
although they have different value of event counter.

    |#1|EVENT|Event name|FIRED|
    |#1|Event body
    |#2|EVENT|Event without name|FIRED|
    |#3|EVENT|Event without name|FIRED|
    |#4|EVENT|Event without name|FIRED|

### Values

With `log_value()` you can log current value of a non-reactive function
or reactive value or expression (inside reactive context).

```{r}
shinyApp(
  ui = fluidPage(
    log_init(),
    selectInput("number", "Select number", choices = c(1, 2, 3), selected = 2)
    ),
  server = function(input, output) {
    set_logging()
    log_value(NROW(mtcars))
    observe({
      log_value(input$number)
    })
 }
)
```

    |#1|VALUE|NROW(mtcars)|FIRED|
    |#1|32
    |#2|VALUE|input$number|FIRED|
    |#2|2

### Outputs

With `log_output()` you can capture the output which is the result of
function evaluation. 
It is especially useful for logging multi-line outputs of irregular structure.

```{r}
shinyApp(
  ui = fluidPage(log_init()),
  server = function(input, output) {
    set_logging()
    log_output(head(mtcars))
    log_output(str(mtcars))
 }
)
```

    |#1|OUTPUT|head(mtcars)|FIRED|
    |#1|                   mpg cyl disp  hp drat    wt  qsec vs am gear carb
    |#1|Mazda RX4         21.0   6  160 110 3.90 2.620 16.46  0  1    4    4
    |#1|Mazda RX4 Wag     21.0   6  160 110 3.90 2.875 17.02  0  1    4    4
    |#1|Datsun 710        22.8   4  108  93 3.85 2.320 18.61  1  1    4    1
    |#1|Hornet 4 Drive    21.4   6  258 110 3.08 3.215 19.44  1  0    3    1
    |#1|Hornet Sportabout 18.7   8  360 175 3.15 3.440 17.02  0  0    3    2
    |#1|Valiant           18.1   6  225 105 2.76 3.460 20.22  1  0    3    1
    |#2|OUTPUT|str(mtcars)|FIRED|
    |#2|'data.frame':	32 obs. of  11 variables:
    |#2| $ mpg : num  21 21 22.8 21.4 18.7 18.1 14.3 24.4 22.8 19.2 ...
    |#2| $ cyl : num  6 6 4 6 8 6 8 4 4 6 ...
    |#2| $ disp: num  160 160 108 258 360 ...
    |#2| $ hp  : num  110 110 93 110 175 105 245 62 95 123 ...
    |#2| $ drat: num  3.9 3.9 3.85 3.08 3.15 2.76 3.21 3.69 3.92 3.92 ...
    |#2| $ wt  : num  2.62 2.88 2.32 3.21 3.44 ...
    |#2| $ qsec: num  16.5 17 18.6 19.4 17 ...
    |#2| $ vs  : num  0 0 1 1 0 1 0 1 1 1 ...
    |#2| $ am  : num  1 1 1 0 0 0 0 0 0 0 ...
    |#2| $ gear: num  4 4 4 3 3 3 3 4 4 4 ...
    |#2| $ carb: num  4 4 1 1 2 1 4 2 2 4 ...

### Unit tests

With `log_test()` you can log a unit test placed inside your
shiny app. 

```{r}
shinyApp(
  ui = fluidPage(log_init()),
  server = function(input, output) {
    set_logging()
    log_test(testthat::expect_true(TRUE))
    log_test(testthat::expect_true(FALSE))
    log_test(testthat::expect_true(TRUE))
 }
)
```

    |#1|TEST|testthat::expect_true(TRUE)|SUCCESS|
    |#2|TEST|testthat::expect_true(FALSE)|ERROR|
    |#2|Error: FALSE isn't true.
    |#3|TEST|testthat::expect_true(TRUE)|SUCCESS|

As you can see, in case a unit test is not successfull, 
error is raised silently and doesn't break the app.

### Diagnostic messages

For classic diagnostic messages you can use wrappers like
`log_message()`, `log_warning()`, and `log_error()`.


```{r}
shinyApp(
  ui = fluidPage(log_init()),
  server = function(input, output) {
    set_logging()
    log_message("Example of a message")
    log_warning("Example of a warning")
    log_error("Example of an error")
 }
)
```

They will log message with appropriate type first, and then 
use `message()`, `warning()`, or `stop()`. 
In case of `log_error()` this can break the running app.

    |#1|MESSAGE|Example of a message|FIRED|
    Example of a message
    |#2|WARNING|Example of a warning|FIRED|
    Warning in log_warning("Example of a warning") : Example of a warning
    |#3|ERROR|Example of an error|FIRED|
    Warning: Error in log_error: Example of an error
      52: stop
      51: log_error [...]
      50: server [#7]
    Error in log_error("Example of an error") : Example of an error


## Timing events

Events can last for some time. 
You can start an instance of an event with `log_started()` 
and log when event finishes with `log_done()`.
This can be used for quick timing an event and save into `secs` parameter 
how many seconds passed since the event started.

```{r}
  shiny::shinyApp(
    ui = shiny::fluidPage(log_init()),
    server = function(input, output) {
      options(digits.secs = 6)
      set_logging(js_console = FALSE)
        log_started("Event 1")
          log_event("Sleeping for 3 secs...")
          Sys.sleep(3)
        log_done("Event 1")
    }
  )
```

    |#1|EVENT|Event 1|STARTED|
    |#2|EVENT|Sleeping for 3 secs...|FIRED|
    |#1|EVENT|Event 1|DONE|
    |#1|PARAMS|list(secs = 3.03)

Please note, that instances of the same event 
have the same value of event counter.

Events can be also nested within other, more general events.


```{r}
  shiny::shinyApp(
    ui = shiny::fluidPage(),
    server = function(input, output) {
      set_logging(js_console = FALSE)
      log_started("Event 1")
        Sys.sleep(1)
          log_started("Event 2")
          Sys.sleep(1)
          log_done("Event 2")
        Sys.sleep(1)
      log_done("Event 1")
    }
  )
```

    |#1|EVENT|Event 1|STARTED|
    |#2|EVENT|Event 2|STARTED|
    |#2|EVENT|Event 2|DONE|
    |#2|PARAMS|list(secs = 1.02)
    |#1|EVENT|Event 1|DONE|
    |#1|PARAMS|list(secs = 3.03)

## Logging events with parameters

There are several ways how you can log events with additional parameters.
You can use:

* event-specific parameters,
* scope-specific parameters,
* app-wide parameters.

### Event-specific parameters

Each event can be logged with a list of additional parameters.

```{r}
shinyApp(
  ui = fluidPage(
    log_init(),
    selectInput("number", "Select number", choices = c(1, 2, 3), selected = 2)
    ),
  server = function(input, output) {
    set_logging()
    log_event("mtcars", params = list(dim = dim(mtcars)))
    observe({
      log_event("number changed", params = list(number = input$number))
    })
 }
)
```

    |#1|EVENT|mtcars|FIRED|
    |#1|PARAMS|list(dim = c(32L, 11L))
    |#2|EVENT|number changed|FIRED|
    |#2|PARAMS|list(number = "2")

### Scope-specific parameters

If you have several events within the same scope 
(for example, within the same observer)
you can set 
some additional parameters that are common to all those events. 
Using `log_params()` allows you to define those parameters only once, 
and not to repeat the same list of parameters with each event.

```{r}
shinyApp(
  ui = fluidPage(
    log_init(),
    selectInput("number", "Select number", choices = c(1, 2, 3), selected = 2)
    ),
  server = function(input, output) {
    set_logging()
    log_event("mtcars", params = list(dim = dim(mtcars)))
    observe({
      log_params(number = input$number)
      log_event("number changed")
      log_event("number changed indeed")
    })
  }
)
```

    |#1|EVENT|mtcars|FIRED|
    |#1|PARAMS|list(dim = c(32L, 11L))
    |#2|EVENT|number changed|FIRED|
    |#2|PARAMS|list(number = "2")
    |#3|EVENT|number changed indeed|FIRED|
    |#3|PARAMS|list(number = "2")

### App-wide parameters

With `set_logging()` you can also set event parameters that are
common for all events in the app.

```{r}
shinyApp(
  ui = fluidPage(
    log_init(),
    selectInput("number", "Select number", choices = c(1, 2, 3), selected = 2)
    ),
  server = function(input, output) {
    set_logging(
      logger_ver = as.character(packageVersion("ShinyEventLogger")),
      build = 123L
    )
    log_event("mtcars", params = list(dim = dim(mtcars)))
    observe({
      log_params(number = input$number)
      log_event("number changed")
      log_event("number changed indeed")
    })
  }
)
```

    |#1|EVENT|mtcars|FIRED|
    |#1|PARAMS|list(dim = c(32L, 11L), build = 123L, logger_ver = "0.0.0.9100")
    |#2|EVENT|number changed|FIRED|
    |#2|PARAMS|list(number = "2", build = 123L, logger_ver = "0.0.0.9100")
    |#3|EVENT|number changed indeed|FIRED|
    |#3|PARAMS|list(number = "2", build = 123L, logger_ver = "0.0.0.9100")








